---
title: "chapter5.Rmd"
author: "Henrik Wejberg"
date: "2022-12-05"
output: html_document
---
# Dimensionality reduction techniques

### Principal component analysis

We will analyse today data from UNDP about variables that make the Human Development Index. The variables we use for analysis today are listed below the distribution analysis.

```{r}
human <- read.table("data/human.csv", header = TRUE, sep = ",", row.names = 1)

# Access GGally
library(GGally)

# visualize the 'human_' variables
ggpairs(human, proportions = "auto")

# Access corrplot
library(corrplot)

# compute the correlation matrix and visualize it with corrplot
cor(human) %>% corrplot()
```
Variable names:      
"GNI" = Gross National Income per capita  
"Life.Exp" = Life expectancy at birth  
"Edu.Exp" = Expected years of schooling   
"Mat.Mor" = Maternal mortality ratio  
"Ado.Birth" = Adolescent birth rate  
"Parli.F" = Percetange of female representatives in parliament  
"Edu2.FM" = Proportion of females with at least secondary education / Proportion of males with at least secondary education  
"Labo.FM" = Proportion of females in the labour force / Proportion of males in the labour force   

The distributions depend significantly on the variable examined. For example, maternal mortality ratio couldn't be quite high in most of the countries. The mode is clearly on the left, where maternal mortality is very low. Most of the variables related to females equality in society, such as Edu2.FM and Labo.FM have modes lower than one, which means that generally females are offered less higher education and participate less to labour force. 

Correlations do not offer any surprises. Edu2.FM has strong positive correlation with expected years of schooling and life expectancy. On the other hand, Edu2.FM has a strong negative correlation with maternal mortality ratio. Not suprisingly, Life expectancy has very strong negative correlation with Maternal mortality ratio and Adolescent birth rate. This is quite obvious, since high life expectancy must mean that not many women or children die in labour. 

We conduct principal component analysis to see which variables explain most of the variance on the sample and are therefore probably the most important ones to keep for further analysis.

```{r}
# perform principal component analysis without scaling
pca_human <- prcomp(human)

s <- summary(pca_human)
print(s)

# rounded percentanges of variance captured by each PC
pca_pr <- round(1*s$importance[2, ], digits = 1)

# print out the percentages of variance
print(pca_pr)

# 
pc_lab <- paste0(names(pca_pr), " (", pca_pr, "%)")

# draw a biplot of the principal component representation and the original variables
biplot(pca_human, choices = 1:2, cex = c(0.4, 0.8), col = c("grey40", "deeppink2"), xlab = pc_lab[1], ylab = pc_lab[2])

```
Result is bad, since the variables are not scaled. There is so much variance because of GNI, that the results are not good. This can be seen from the plot. Principal component 1 explains 100 % of variance, because the variance of GNI is much more higher than other variables. This is due to the fact that the range of GNI is much higher than other variables, which are share variables or are between human death ages in different countries. We need to scale the variables in order to have better results.

```{r}

# standardize the variables
human_std <- scale(human)

# print out summaries of the standardized variables
summary(human_std)

# perform principal component analysis with scaled variables
pca_human_std <- prcomp(human_std)

s2 <- summary(pca_human_std)
print(s2)

# rounded percentanges of variance captured by each PC
pca_pr_2 <- round(1*s2$importance[2, ], digits = 2)

# print out the percentages of variance
print(pca_pr_2)

# 
pc_lab_2 <- paste0(names(pca_pr_2), " (", pca_pr_2, "%)")

# draw a biplot of the principal component representation and the original variables
biplot(pca_human_std, choices = 1:2, cex = c(0.4, 0.8), col = c("grey40", "deeppink2"), xlab = pc_lab_2[1], ylab = pc_lab_2[2], sub = "GNI, Edu.Exp, Life.Exp and Edu2.FM are highly correlated with each other and negatively correlated with Maternal Mortality and Adolescent Birth.") 

```
Now the plot and results look much better! It seems that PC1 accounts for over half the variance. In this horizontal axis, there are 6 variables which are significantly correlated with each other positively or negatively. I think that this component relates to generally the education level and GNI of the country. Good GNI requires good education and a sufficiently large labor force, which requires that females also work outside of home. If people have good education, high income and therefore high life expectancy on birth, it is highly unlikely that there is maternal mortality to significant degree or adolescent births. The countries in the right side of this PC1 seem to be African countries and to the left mostly "Western" countries. 

Second PC relates to females participation in parliaments and work outside of home. Nordic countries and Northern European countries rank highly on these, while Middle East countries have significantly different results. If we would like to have less variables or indexes, we could have a variable related maybe to GNI and to life expectancy and education and another one for females participation in work and democratic processes. 

### Multiple correspondence analysis

```{r, echo=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(FactoMineR)
tea <- read.csv("https://raw.githubusercontent.com/KimmoVehkalahti/Helsinki-Open-Data-Science/master/datasets/tea.csv", stringsAsFactors = TRUE)
```


```{r}
# column names to keep in the dataset
keep_columns <- c("Tea", "How", "how", "sugar", "where", "lunch")

# select the 'keep_columns' to create a new dataset
tea_time <- dplyr::select(tea, all_of(keep_columns))

# look at the summaries and structure of the data
summary(tea_time)
str(tea_time)

# visualize the dataset
library(ggplot2)
pivot_longer(tea_time, cols = everything()) %>% 
  ggplot(aes(value)) + facet_wrap("name", scales = "free") +
  geom_bar() + theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8))
```

The data is related to how tea is enjoyed. There is a nice plot which shows the counts of different factor values. People consume tea mostly in chain stores. Sugar use is evenly distributed. Earl Grey is the mode of different teas. Many people drink tea outside of lunch. 


```{r}
# multiple correspondence analysis
mca <- MCA(tea_time, graph = FALSE)

# summary of the model
summary(mca)

# visualize MCA
plot(mca, invisible=c("ind"), graph.type = "classic", habillage = "quali")

```
Tea shop and unpackaged tea are far away from the origin, which means that they are not consumed very often. This can be seen from Dim 1. When these are coupled, they are also far away from origin in Dim 2. It is also rare to drink tea in itself or with lemon or milk. I do not do with what you drink the tea if you do not have any of these but some other stuff put in. Mostly tea is consumed with sugar, not after lunch, alone and the tea varieties are Earl Grey or Black tea. 
